---
title: "Tutorial on Plotting"
author: "Nina Mainusch, Marc Viladrich"
date: "28 11 2019"
output: html_document
---

## Organizational issues: 
- Please merge to groups of at least 2 people

- READ the instructions of the homeworks again: 
-> Name your file correctly (i.e. **IDA_HW4-Group-XYZ.zip**)
-> Don't suppress your code in the knitted file. We need to see your code AND the output created by your code!! (Let "echo" be TRUE, don't set it to FALSE)
-> if you upload multiple versions for one homework, delete the old ones!


## Common mistakes in homework 3

### Exercise 4: Toying with mean and median

Give an example of a metric vector and a single number such that adding the number to the vector does not change the median at all, but does change the mean drammatically. (Use R for the calculation of mean and median, so that there are no lingering doubts about how exactly to compute the median in case of ties etc.)

```{r}
# Sample solution
metric_vector <- c(1,2,3,3,4,5)

m <- mean(metric_vector)
med <- median(metric_vector)

metric_vector_plus_one <- c(metric_vector, 30000)

m_1 <- mean(metric_vector_plus_one)
med_1 <- median(metric_vector_plus_one)

# median not changed
# mean dramatically changed
tibble(median = c(med, med_1), mean = c(m, m_1))
```




## Plotting

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```


### Research question

*Do cars with big engines use more fuel than cars with small engines?* 
You probably already have an answer, but try to make your answer precise. What does the relationship between engine size and fuel efficiency look like? Is it positive? Negative? Linear? Nonlinear?

Let's investigate this!


#### Our data

```{r}
glimpse(mpg)
```


Check out these variables of our data set:

- `displ`, a car’s engine size in litres against

- `hwy`, a car’s fuel efficiency on the highway, in miles per gallon (mpg). A car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance.

```{r}
# plot displ against hwy
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


```{r}
# change axis labels for better understanding what is going on
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  xlab("engine size in litres") +
  ylab("fuel efficiency on the highway (mpg)")
```



#### A graphing template

```{r, eval = FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```


### Exercises

1. Run `ggplot(data = mpg)`. What do you see?
```{r}

```

2. Make a scatterplot of `hwy` vs `cyl`

```{r}

```

3. What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r}

```



### Aesthetic mappings

You can map the colors of your points to the `class` variable to reveal the class of each car.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

There are many more aesthetic mappings, like `alpha` and `shape`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

You can also set the aesthetic properties of your geom manually. For example, we can make all of the points in our plot blue:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")

```
Note that to set an aesthetic manually it goes outside of the `aes()` part.




#### Exercise

1. Make a scatterplot of `displ` against `hwy` which also shows the different categories for cylinders and drive system. Also, make all points red.

```{r}


```


### Labeling

It is important that you always label your plots to make them understandable to people who are not familiar with your work. You can use the `labs` component for this.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = drv, alpha = cyl), color = "red") + 
  labs(
    title = "Highway fuel efficiency goes down with increasing engine size",   # main finding
    subtitle = "The correlation is rather strong",      # additional info on the main finding
    x = "engine size in litres",                        # x-axis label
    y = "fuel efficiency on the highway",               # y-axis label
    tag = "A",                                          # often used to distinguish related (sub-)plots
    caption = "data source: tidyverse's mpg dataset",   # often used to indicate the data source
    alpha = "No. of Cylinders",                         # you can customise the legend labels of the plot
    shape = "Drive System"                              # you can customise the legend labels of the plot
  )
```



### Geom Smooth

Fits a regression line onto the data. The grey area around the line represents the confidence interval (set to 0.95 by default).
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r}
# Display multiple geoms in the same plot
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

# Or to avoid duplications
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

If you place mappings in a geom function, ggplot2 will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

You can use the same idea to specify different data for each layer. Here, our smooth line displays just a subset of the mpg dataset, the subcompact cars. The local `data` argument in geom_smooth() overrides the global `data` argument in ggplot() for that layer only.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```



```{r}
# Display legend what colours mean
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, colour = drv))

# Do not display legend
 ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, colour = drv), show.legend = FALSE)
```




### Statistical transformations

#### Bar charts
Bar charts are a nice tool to visualise data across several categories. There are two ways of creating bar charts with ggplot, `geom_bar` and `geom_col`.



##### geom_bar
We use `geom_bar` when we want to visualise frequencies in our data, i.e. the number of observations belonging to a certain category.
```{r}
# Returns the number of cars in each car class
ggplot(data = mpg) +
  geom_bar(aes(x = class))
```

```{r}
# You can also just look at a subset of your data
ggplot(data = mpg, aes(x = class)) +
  geom_bar() +
  geom_bar(data = filter(mpg, class == "subcompact"), fill = "blue")
```

```{r}
# You can also include other nominal variables in your bar chart (a.k.a. "stacking")
ggplot(data = mpg, aes(x = class)) +
  geom_bar() +
  geom_bar(data = filter(mpg, class == "subcompact"), aes(fill = drv))

# Without filtering
ggplot(data = mpg, aes(x = class)) +
  geom_bar(aes(fill = drv))
```




##### geom_col
We use `geom_col` when we want to visualise the actual value of a certain variable across different categories.

Let's take a look at the dataset `txhousing` which provides some data on the housing market in Texas from 2000 to 2015.
```{r}
txhousing %>% glimpse
```

```{r, eval = FALSE}
# Won't work, we need to specify the second variable
ggplot(data = txhousing) +
  geom_col(aes(x = year))
```

```{r}
ggplot(data = txhousing) +
  geom_col(aes(x = as_factor(year), y = sales)) +     # treat year as a nominal variable
  scale_y_continuous(labels = scales::comma) +        # change the labels on the y-axis to make them more readable
  xlab("year")                                        # rename x-axis
```


#### Exercise

1. Have a look at this dataset of demographic data from a couple states in the Midwest of the United States.

```{r}
midwest %>% glimpse
```

Let's say I want to count the number of counties in each state using one bar chart and count the number of people in each state using another bar chart. Do I use `geom_bar` or `geom_col`?

```{r}
# YOUR ANSWER HERE

```



#### Boxplots

Boxplots are a more visually pleasing way of representing the information given in the *five value summary*. One big advantage of boxplots is that you can easily see **where your data is centered and how much it spreads**. The larger the box, the more your data spreads (i.e. the higher the variance). <br />
The bottom of the box marks the 1st *quartile*, the top of the box marks the 3rd *quartile.* The bold line inbetween marks the *median* (a.k.a. the 2nd *quartile*). <br />
The lines coming out of the box are called *whiskers*. The bottom whisker extends from the bottom of the box to the smallest value that is still **within 1.5 * IQR**, where *IQR* is the interquartile range (i.e. the difference between the 3rd and the 1st quartile). The top whisker extends from the top of the box to the largest value that is still within 1.5 * IQR. <br />
Any value that falls outside the whiskers is marked as an *outlier* and plotted separately. <br />
Thus, the minimum and maximum values may either be an outlier or the end of the whiskers.

```{r}
# Shows the distribution of highway fuel efficiency for each car class
ggplot(data = mpg) +
  geom_boxplot(
    aes(x = class, y = hwy),
    outlier.colour = "red",      # customise outlier appearance
    outlier.shape = NA           # hide outliers, does NOT remove them from the data
  )
```



#### Exercise

1. Make a boxplot that shows the distribution of the highway fuel efficiency depending on the number of cylinders. Keep in mind that boxplots visualise data across different categories.

```{r}
# YOUR CODE HERE
```



##### Plot-grid

Sometimes it is helpful to show multiple plots at once side by side. The package `cowplot` can help us do that with its `plot-grid` function.

```{r}
library(cowplot)
```

```{r}
# Plot fuel efficiency and engine size
mapping = aes(x = displ, y = hwy)
# Define one y-scale such that plots are easier to compare
y_scale <- scale_y_continuous(limits = c(10, 45))


mileage_vs_engine_1999 <- ggplot(data = filter(mpg, year == 1999), mapping = mapping) +
  geom_point() +
  geom_smooth() +
  y_scale

mileage_vs_engine_2008 <- ggplot(data = filter(mpg, year == 2008), mapping = mapping) +
  geom_point() +
  geom_smooth() +
  y_scale

plot_grid(mileage_vs_engine_1999, mileage_vs_engine_2008)
```

#### Exercise

1. Use `cowplot`'s `plot_grid` function and the dataset `txhousing` to compare the evolution of the median house price along the years in the cities Killeen-Fort Hood, Dallas, Texarkana and Austin. Give all plots a title to keep everything organised.

```{r}
# Use this scale to compare the plots more easily (simply add it as another component to your ggplot call)
y_scale <- scale_y_continuous(limits = c(8 * 10^4, 2.5 * 10^5) , labels = scales::comma)
```

```{r}
# YOUR CODE HERE
```

